Data Munging

### 2020-02-25 scripts/dataMunge.Rmd
fixing missing nitrate values

## Calculate dissolved oxygen and stratification indices from profile data
Water temperature and dissolved oxygen profiles were measured at the index site in each waterbody.  A variety of metrics describing the degree of thermal stratification can be extracted from these data.  Webb et al. used the Brunt–Väisälä buoyancy frequency as an indicator of thermal stratification strength.  We will do the same here.

@@

Webb et al. used the 'surface' and 'bottom' water DO percent saturation and temperature as predictors, but didn't give more details on how these indices were calculated.  We will do something similar and define the 'shallowest' and 'deepest' measurements as 'surface' and 'bottom', respectively, but will use mg.l for dissolved oxygen.
```{r}
# o2 surface (s) and bottom (b)
o2.s.b <- 
temp.o2.2 %>% 
  group_by(SITE_ID, VISIT_NO) %>%
  slice(c(which.min(DEPTH), which.max(DEPTH))) %>% # min and max depth by group
  arrange(SITE_ID, VISIT_NO, DEPTH) %>%
    mutate(DEPTH = ifelse(row_number() == 1,
                        "surface",
                        "bottom")) %>%
  pivot_wider(id_cols = c(SITE_ID, VISIT_NO), 
              names_from = DEPTH, 
              values_from = c(TEMPERATURE.M, OXYGEN.M)) %>%
  rename(temp.surf = TEMPERATURE.M_surface,
         temp.bottom = TEMPERATURE.M_bottom,
         o2.surf = OXYGEN.M_surface,
         o2.bottom = OXYGEN.M_bottom)

@@

Write out final .RData object.
```{r}
save(dg, file = paste0(localPath, 
                       "/Environmental Protection Agency (EPA)/",
                       "ORD NLA17 Dissolved Gas - Documents/",
                       "inputData/dg.", Sys.Date(), ".RData"))
```

>>>>

## Calculate buoyancy frequency
Water temperature and dissolved oxygen profiles were measured at the index site in each waterbody.  A variety of metrics describing the degree of thermal stratification can be extracted from these data.  Webb et al. used the Brunt–Väisälä buoyancy frequency as an indicator of thermal stratification strength.  We will do the same here.

@@

## Calculate surface and bottom DO/temp
Webb et al. used the 'surface' and 'bottom' water DO percent saturation and temperature as predictors, but didn't give more details on how these indices were calculated.  We will do something similar and define the 'shallowest' and 'deepest' measurements as 'surface' and 'bottom', respectively, but will use mg.l for dissolved oxygen.
```{r}
# o2 surface (s) and bottom (b)
o2.s.b <- 
temp.o2.2 %>% 
  group_by(SITE_ID, VISIT_NO) %>%
  slice(c(which.min(DEPTH), which.max(DEPTH))) %>% # min and max depth by group
  arrange(SITE_ID, VISIT_NO, DEPTH) %>%
    mutate(DEPTH = ifelse(row_number() == 1,
                        "surface",
                        "bottom")) %>%
  pivot_wider(id_cols = c(SITE_ID, VISIT_NO), 
              names_from = DEPTH, 
              values_from = c(TEMPERATURE.M, OXYGEN.M)) %>%
  rename(temp.surf = TEMPERATURE.M_surface,
         temp.bottom = TEMPERATURE.M_bottom,
         o2.surf = OXYGEN.M_surface,
         o2.bottom = OXYGEN.M_bottom)

@@

## Fix chemistry
The file contains `r dg %>% dplyr::select(nitrate.n.result) %>% filter(is.na(nitrate.n.result)) %>% summarize(na.no3 = nrow(.)) %>% pull()` missing nitrate values.  Inspection of raw data sheet indicates that:
* North Dakota reported NITRATE_NITRITE_N_RESULT rather than NITRATE.N.RESULT.  Most of the ND sites were 'non detect (ND)' with a detection limit of 0.03 mg /L.
* WI sites were ND with a detection limit of 0.019.

Will replace these values with NITRATE_NITRITE_N_RESULT when available, else 1/2 detection limit.

``` {r}
dg <- dg %>% 
  mutate(nitrate.n.result = replace(nitrate.n.result, 
                                    grepl("WI", site.id) & is.na(nitrate.n.result), # WI na sites
                                    0.019/2), #1/2 detection limit
         # ND sites without nitrate.n and nitrate.nitrite
         nitrate.n.result = replace(nitrate.n.result, 
                                    grepl("ND", site.id) & 
                                      is.na(nitrate.n.result) & 
                                      is.na(nitrate.nitrite.n.result), 
                                    0.03/2), # 1/2 detection limit
# ND sites with a nitrate.nitrite.n.result
         nitrate.n.result = replace(nitrate.n.result, 
                                    grepl("ND", site.id) & 
                                      is.na(nitrate.n.result) & 
                                      !is.na(nitrate.nitrite.n.result), # ND na sites
                                    0.03)) # two sites reporting 0.03 mg/L
```

The file contains `r dg %>% filter(is.na(chla.result.vol)) %>% summarize(na.chl = nrow(.)) %>% pull()` missing chlorophyll values.  Inspection reveals that these are all from North Dakota sites.  It looks like the lab failed to record the volume filtered, so we can't calculate ug/L chl from these sites.  These will remain missing.

## Write out data
Write out final .RData object.
```{r}
save(dg, file = paste0(localPath, 
                       "/Environmental Protection Agency (EPA)/",
                       "ORD NLA17 Dissolved Gas - Documents/",
                       "inputData/dg.", Sys.Date(), ".RData"))
```
